---
title: "Reproducible Research: Peer Assessment 1"
author: "Jeremy Eade"
date: "May 28, 2016"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r readdata}
library(dplyr)

options(scipen=999,digits=2)

df <- read.csv("activity.csv")
df <- tbl_df(df)
str(df)
```

   
      
      
## What is mean total number of steps taken per day?

<span style="color:blue">1. Calculate the total number of steps taken per day</span>

```{r total_steps_day}
by_day <- df %>% group_by(date) %>% summarize(sum=sum(steps,na.rm=TRUE))

head(by_day)
summary(by_day$sum,na.rm=TRUE)

```


<span style="color:blue">2. Make a histogram of the total number of steps taken each day</span>


```{r total_steps}

steps_mean <- mean(by_day$sum,na.rm = TRUE)
steps_median <- median(by_day$sum, na.rm=TRUE)
hist(by_day$sum,breaks = seq(from=0,to=24000,by=2000),col="mistyrose1",
     main="Histogram of the total number of steps taken each day",
     xlab = "Total Number of Steps per Day")

abline(v = steps_mean,
       col = "lightblue",
       lwd = 5)

abline(v = steps_median,
         col = "magenta4",
         lwd = 1)

legend(x = "topright",
       c("Mean", "Median"),
       col = c("lightblue", "magenta4"),
       lwd = c(5, 1))

```


<span style="color:blue">3. Calculate and report the mean and median of the total number of steps taken per day</span>

The mean of the total number of steps taken per day is **`r steps_mean`** and the median is **`r steps_median`**.




## What is the average daily activity pattern?


<span style="color:blue">1. Make a time series plot (i.e. ğšğš¢ğš™ğš = "ğš•") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)</span>


```{r interval_mean}
by_interval<- df %>% group_by(interval) %>% summarize(avg=mean(steps,na.rm=TRUE))
max_interval <- filter(by_interval,avg==max(by_interval$avg))$interval
max_steps <- max(by_interval$avg)

str(by_interval)

with(by_interval,plot(interval,avg,type="l",col="blue",
                      main="Average Steps Taken by Interval",xlab="Interval",ylab="Average Number of Steps"))
abline(v = max_interval,
        col = "magenta4",
        lwd = 1)
 
legend(x = "topright",
        c("Max Steps"),
        col = c("magenta4"),
        lwd = c(1))

```


<span style="color:blue">2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?</span>

The answer is interval=**`r max_interval`** with **`r max_steps`** steps.


   
   
## Imputing missing values


<span style="color:blue">1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with ğ™½ğ™°s)
</span>

```{r total_na}
na_rows <- sum(is.na(df$steps))
total_rows <- NROW(df$steps)
```

The total number of rows with NAs is **`r na_rows`** out of a total of **`r total_rows`** rows.


<span style="color:blue">2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 
For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.</span>

Strategy - we will use the mean for that 5-minute interval

<span style="color:blue">3. Create a new dataset that is equal to the original dataset but with the missing data filled in.</span>

```{r imputedata}
join <- inner_join(df,by_interval)
join <- mutate(join,steps2=ifelse(is.na(steps),avg,steps))
sum(is.na(join$steps))
sum(is.na(join$steps2))
summary(join$steps)
summary(join$steps2)
```

<span style="color:blue">4. Make a histogram of the total number of steps taken each day</span>

```{r histogram_steps_imputed}
by_day2 <- join %>% group_by(date) %>% summarize(sum=sum(steps2))

summary(by_day2$sum)

steps2_mean <- mean(by_day2$sum)
steps2_median <- median(by_day2$sum)
hist(by_day2$sum,breaks = seq(from=0,to=24000,by=2000), col="mistyrose1",
     main="Histogram of the total number of steps taken each day (Imputed)",
     xlab = "Total Number of Steps per Day")

abline(v = steps2_mean,
       col = "lightblue",
       lwd = 5)

abline(v = steps2_median,
       col = "magenta4",
       lwd = 1)

legend(x = "topright",
       c("Mean", "Median"),
       col = c("lightblue", "magenta4"),
       lwd = c(5, 1))
```
   
       
<span style="color:blue">5. Calculate and report the mean and median total number of steps taken per day. </span>

The mean of the total number of steps taken per day is **`r steps2_mean`** and the median is **`r steps2_median`**.

<span style="color:blue">6. Do these values differ from the estimates from the first part of the assignment? </span>


```{r diff}
diff_mean <- abs(steps_mean-steps2_mean)
diff_median <- abs(steps_median-steps2_median)
```

Yes!  There are surely differences:   

- the difference between the mean before and after imputing missing NAs is **`r diff_mean`**.   
- the difference between the median before and after imputing missing NAs is **`r diff_median`**.


<span style="color:blue">7. What is the impact of imputing missing data on the estimates of the total daily number of steps?</span>

```{r estimated_total_steps}
total_steps <- sum(join$steps,na.rm=TRUE)
total_steps_imputed <- sum(join$steps2)
```

The total number of steps with missing data is **`r total_steps`**.   
The total number of steps with imputed data is **`r total_steps_imputed`**.   
Thus imputing the missing data causes the total number of steps to increase.   





## Are there differences in activity patterns between weekdays and weekends?


<span style="color:blue">1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.</span>


```{r dayOfweek}
library(lubridate)
join <- mutate(join,dayOfWeek = ifelse(weekdays(ymd(date)) %in% c("Sunday","Saturday"),"weekend","weekday"))
by_dayOfWeek <- join %>% group_by(dayOfWeek,interval) %>% summarize(avg=mean(steps2))
str(by_dayOfWeek)
table(by_dayOfWeek$dayOfWeek)
```




<span style="color:blue">2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). </span>


```{r time_series plot}
library(lattice)
xyplot(avg~interval|dayOfWeek,data=by_dayOfWeek,type="l",layout=c(1,2),xlab="Interval",ylab="Average Number of Steps")
```
